---
title: "36-671 Final Project"
author: "Ellie Huang, Nikita Jayaprakash, Annie Wang, Aishani Pradhan, Steve Wong"
date: "2026-04-19"
format: pdf
---

## Research Question

How do regional gasoline prices co-move over time, and what are the lead-lag relationships between regions and between crude oil prices and retail gasoline prices?

## 1. Preparing Time Series

```{r}
library(here)
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(GGally)
library(vars)
library(forecast)
library(lubridate)

file_path <- here("data", "PET_PRI_GND_A_EPM0_PTE_DPGAL_W.xls")
brent_path <- here("data", "brent_crude_weekly_1993_present.csv")
```

```{r}
# Load weekly gasoline data
# skip = 2 because row 1 = title, row 2 = source key, row 3 = actual variable names
us_data <- read_excel(file_path, sheet = "Data 1", skip = 2) |>
  rename(Date = 1, US_All_Grades = 2) |>
  mutate(Date = as.Date(Date), US_All_Grades = as.numeric(US_All_Grades)) |>
  arrange(Date)

region_data <- read_excel(file_path, sheet = "Data 2", skip = 2) |>
  rename(
    Date = 1,
    East_Coast = 2,
    New_England = 3,
    Central_Atlantic = 4,
    Lower_Atlantic = 5,
    Midwest = 6,
    Gulf_Coast = 7,
    Rocky_Mountain = 8,
    West_Coast = 9,
    West_Coast_ex_CA = 10
  ) |>
  mutate(Date = as.Date(Date)) |>
  mutate(across(-Date, as.numeric)) |>
  arrange(Date)

# Join national + five regional series, drop any incomplete weeks
gas_complete <- us_data |>
  left_join(
    region_data |> dplyr::select(Date, East_Coast, Midwest, Gulf_Coast, Rocky_Mountain, West_Coast),
    by = "Date"
  ) |>
  drop_na()

cat("Weekly gasoline date range:", min(gas_complete$Date), "to", max(gas_complete$Date), "\n")
cat("Observations:", nrow(gas_complete), "\n")

# Load weekly Brent crude oil data
crude_data <- read_csv(brent_path) |>
  rename(Brent = Brent_Price_USD) |>
  mutate(Date = as.Date(Date), Brent = as.numeric(Brent)) |>
  arrange(Date)

cat("Crude oil date range:", min(crude_data$Date), "to", max(crude_data$Date), "\n")
```

```{r}
# Convert all series to weekly ts objects
# Get week number via %U (Sunday start); +1 to make it start from 1
start_vec <- c(
  as.numeric(format(min(gas_complete$Date), "%Y")),
  as.numeric(format(min(gas_complete$Date), "%U")) + 1
)

us_ts <- ts(gas_complete$US_All_Grades, start = start_vec, frequency = 52)
east_ts <- ts(gas_complete$East_Coast, start = start_vec, frequency = 52)
midwest_ts <- ts(gas_complete$Midwest, start = start_vec, frequency = 52)
gulf_ts <- ts(gas_complete$Gulf_Coast, start = start_vec, frequency = 52)
rocky_ts <- ts(gas_complete$Rocky_Mountain, start = start_vec, frequency = 52)
west_ts <- ts(gas_complete$West_Coast, start = start_vec, frequency = 52)

# Build multivariate ts for VAR analysis
regions_ts <- gas_complete |>
  dplyr::select(East_Coast, Midwest, Gulf_Coast, Rocky_Mountain, West_Coast) |>
  as.matrix() |>
  ts(start = start_vec, frequency = 52)

colnames(regions_ts) <- c("East_Coast", "Midwest", "Gulf_Coast", "Rocky_Mountain", "West_Coast")
```

```{r}
# Both series are weekly, join directly on date
gas_crude <- gas_complete |>
  inner_join(crude_data, by = "Date") |>
  drop_na()

cat("Gas price-crude oil overlapping date range:", min(gas_crude$Date), "to", max(gas_crude$Date), "\n")
cat("Observations:", nrow(gas_crude), "\n")

# Shared start for weekly ts
start_crude <- c(
  as.numeric(format(min(gas_crude$Date), "%Y")),
  as.numeric(format(min(gas_crude$Date), "%U")) + 1
)

gas_ts_crude <- ts(gas_crude$US_All_Grades, start = start_crude, frequency = 52)
crude_ts <- ts(gas_crude$Brent, start = start_crude, frequency = 52)
```

## 2. Exploratory Data Analysis

```{r}
# Time series plots: national, regional, and crude oil
plot(us_ts,
     main = "U.S. All Grades Gasoline Prices (Weekly)",
     xlab = "Year", ylab = "Dollars per Gallon")

ts.plot(east_ts, midwest_ts, gulf_ts, rocky_ts, west_ts,
        col = 1:5, lty = 1,
        main = "Regional Weekly Gasoline Prices",
        xlab = "Year", ylab = "Dollars per Gallon")
legend("topleft",
       legend = c("East Coast", "Midwest", "Gulf Coast", "Rocky Mountain", "West Coast"),
       col = 1:5, lty = 1, bty = "n")

# Standardize for comparison since crude oil and gas prices are on diff scales/units
crude_scaled <- ts(
  as.numeric(scale(as.numeric(crude_ts))),
  start = start(crude_ts),
  frequency = frequency(crude_ts)
)

gas_scaled <- ts(
  as.numeric(scale(as.numeric(gas_ts_crude))),
  start = start(gas_ts_crude),
  frequency = frequency(gas_ts_crude)
)

ts.plot(crude_scaled, gas_scaled,
        col = c("steelblue", "firebrick"), lty = 1,
        main = "Standardized Brent Crude Oil vs. U.S. Gasoline Prices (Weekly)",
        xlab = "Year", ylab = "Standardized Value")
legend("topleft",
       legend = c("Brent Crude Oil", "U.S. Gasoline"),
       col = c("steelblue", "firebrick"), lty = 1, bty = "n")
```

The national gasoline series shows long swings over time rather than fluctuating around a constant level, suggesting strong persistence and possible nonstationarity in the level series. There seems to be a general upward movement over the sample. The plots also suggest changing variability over time, with larger swings during some periods, especially around major shocks, which may indicate heteroskedasticity.

The regional plot shows that all five regions follow the same broad pattern, rising and falling during the same major episodes.

At the same time, the regions are not exactly the same. The West Coast is generally higher than the other regions and appears more volatile, especially in later years. Rocky Mountain also shows somewhat greater variability at certain points, while the Gulf Coast and Midwest tend to track the national pattern more closely and at lower price levels. So the data show both strong shared movement and meaningful regional differences.

Comparing Brent crude to U.S. gasoline, the two series move very closely together over time. This suggests that crude oil is an important driver of gasoline prices, although there are other factors like refining, transportation, taxes, regional conditions, etc.

##### Question: Ask prof if we need to include these scatterplots too, or just the lagged scatters.

```{r}
# Pairwise scatterplots to check contemporaneous correlations before looking at lags
gas_complete |>
  dplyr::select(East_Coast, Midwest, Gulf_Coast, Rocky_Mountain, West_Coast) |>
  ggpairs(
    title = "Pairwise Regional Gasoline Prices",
    lower = list(continuous = wrap("points", alpha = 0.2, size = 0.5)),
    upper = list(continuous = wrap("cor", size = 3))
  )
```

The pairwise scatterplots indicate strong positive contemporaneous relationships among weekly gasoline prices across all regions. For every regional pair, the points form a clear upward-sloping pattern, suggesting that when prices are relatively high in one region, they also tend to be relatively high in the other during the same week. This is consistent with the very high pairwise correlations, which range from 0.945 to 0.996. Several pairs, such as East Coast and Midwest, fall especially tightly along a line, indicating very strong same-week linear association. Pairs involving the West Coast appear somewhat more dispersed by comparison, though the relationship remains strongly positive. However, because these plots compare observations from the same week only, they do not provide evidence about lead-lag structure or whether price changes in one region tend to precede changes in another.

This shows that the regions are clearly not independent; therefore it is reasonable to study them as a connected multivariate time series. Having established strong contemporaneous correlation, we can next investigate whether some regions tend to lead or lag others over time. This provides a natural motivation for using lag-based analysis and vector autoregressive models.

## 3. Crude Oil Prices vs US Gasoline Price Analyses

### (i) Converting to Stationary Series

Both crude oil and gasoline prices exhibit strong trends over time, so we first transform the series to make them approximately stationary before analyzing their lag relationships.

```{r}
# Log difference 
dGas <- diff(log(gas_ts_crude))
dBrent <- diff(log(crude_ts))

#par(mfrow = c(3,2))
plot(dGas, main = "Differenced Log U.S. Gasoline")
plot(dBrent, main = "Differenced Log Brent Crude")
acf(dGas, lag.max = 20, main = "ACF: Differenced Log Gasoline")
acf(dBrent, lag.max = 20, main = "ACF: Differenced Log Brent")
pacf(dGas, lag.max = 20, main = "PACF: Differenced Log Gasoline")
pacf(dBrent, lag.max = 20, main = "PACF: Differenced Log Brent")
par(mfrow = c(1,1))
```

After taking first differences of the log prices, both series fluctuate around a stable mean with rapidly decaying ACFs. The PACFs show one dominant spike at lag 1 followed by little remaining dependence, which is consistent with a low order AR process and suggests that the transformed series are sufficiently stationary for CCF and VAR analysis.

### (ii) CCF and LR to Find Lag Dependence

```{r}
# Negative lag means Brent leads gasoline
ccf(dBrent, dGas,
    lag.max = 12,
    main = "CCF: Brent vs U.S. Gasoline (Differenced Log Prices)")
```

The cross-correlation of the differenced log prices is largest at lag 0, indicating that gasoline and crude oil prices move together strongly within the same week. There are also significant positive correlations at lags of approximately -1 to -3 weeks, suggesting that changes in Brent crude prices continue to pass through into gasoline prices over the following several weeks.

##### Question: how does the instantaneous correlation impact our analysis?

```{r}
oil_gas_lag <- ts.intersect(
  gas = dGas,
  brentL1 = stats::lag(dBrent, -1),
  brentL2 = stats::lag(dBrent, -2),
  brentL3 = stats::lag(dBrent, -3),
  dframe = TRUE
)

# Scatterplots

par(mfrow = c(1,3))

plot(oil_gas_lag$brentL1, oil_gas_lag$gas,
     xlab = "Brent change 1 week earlier",
     ylab = "Current gasoline change",
     main = "Lag 1")
abline(lm(gas ~ brentL1, data = oil_gas_lag), col = "red")

plot(oil_gas_lag$brentL2, oil_gas_lag$gas,
     xlab = "Brent change 2 weeks earlier",
     ylab = "Current gasoline change",
     main = "Lag 2")
abline(lm(gas ~ brentL2, data = oil_gas_lag), col = "red")

plot(oil_gas_lag$brentL3, oil_gas_lag$gas,
     xlab = "Brent change 3 weeks earlier",
     ylab = "Current gasoline change",
     main = "Lag 3")
abline(lm(gas ~ brentL3, data = oil_gas_lag), col = "red")

par(mfrow = c(1,1))
```

```{r}
# Check if lagged valued of crude oil week 1-3 are signficant in predicting gas prices
fit_oil <- lm(gas ~ brentL1 + brentL2 + brentL3,
              data = oil_gas_lag)

summary(fit_oil)
```

The lagged scatterplots and regression reinforce the CCF results. The relationship between crude oil and gasoline changes is strongest after one week, with weaker but still positive associations after two and three weeks. This supports the conclusion that gasoline prices adjust gradually over several weeks following a change in crude oil prices.

### (iii) VAR(2) Model

```{r}
library(vars)

# Combine the stationary series
x_oil <- cbind(
  Gasoline = dGas,
  Brent = dBrent
)

# choose lag length
VARselect(x_oil, lag.max = 8, type = "none")
```

```{r}
fit_var_oil <- VAR(x_oil, p = 2, type = "none")
summary(fit_var_oil)
```

##### Questions:

-   Which VAR model to use in terms of AIC, HQ, SC, FPE? SC with parameter = 2 because simplest?

    ```         
    AIC(n)  HQ(n)  SC(n) FPE(n) 
    ```

-   why would the effect of gasoline (l2) be stronger on crude oil than crude oil on gasoline (estimates)? remaining correlation? are the estimates even comparable?

### (iv) Diagnostics

```{r}
# Fitted values and residuals from VAR(2)
fitted_vals <- fitted(fit_var_oil)
res <- resid(fit_var_oil)
```

```{r}
# Observed and fitted gasoline changes
obs_gas <- window(x_oil[, "Gasoline"], start = time(x_oil)[3])
fit_gas <- ts(fitted_vals[, "Gasoline"],
              start = start(obs_gas),
              frequency = frequency(obs_gas))

plot(obs_gas, type = "l", col = "black",
     main = "Observed vs Fitted: Gasoline",
     ylab = "Differenced Log Gasoline", xlab = "Time")
lines(fit_gas, col = "red")
legend("topright",
       legend = c("Observed", "Fitted"),
       col = c("black", "red"), lwd = 1)
```

```{r}
plot(obs_gas, fit_gas,
     pch = 16,
     xlab = "Observed",
     ylab = "Fitted",
     main = "Observed vs Fitted: Gasoline")
legend("topleft", 
       legend = paste("Correlation:", 
                      round(cor(obs_gas, fit_gas), 3)),
       bty = "n")
abline(0, 1, col = "red", lwd = 2)
```

```{r}
## GAS
# top 10 residuals
res_gas <- res[,1]
time <- time(dGas)[-c(1,2)]
top_resid_gas <- order(abs(res_gas), decreasing = TRUE)[1:10]
# corresponding times in date format
top_resid_gas_time <- time[top_resid_gas] |>
  date_decimal() |>
  as.Date()
# in tabular format
tibble::tibble(
  `Time (Gas)` = top_resid_gas_time,
  `Top Residuals (Gas)` = res_gas[top_resid_gas])
```

```{r}
## BRENT
# top 10 residuals
res_brent <- res[,2]
time <- time(dBrent)[-c(1,2)]
top_resid_brent <- order(abs(res_brent), decreasing = TRUE)[1:10]
# corresponding times in date format
top_resid_brent_time <- time[top_resid_brent] |>
  date_decimal() |>
  as.Date()
# in tabular format
tibble::tibble(
  `Time (Brent)` = top_resid_brent_time,
  `Top Residuals (Brent)` = res_brent[top_resid_brent]
)
```

```{r}
# observed and fitted Brent changes
obs_brent <- window(x_oil[, "Brent"], start = time(x_oil)[3])
fit_brent <- ts(fitted_vals[, "Brent"],
                start = start(obs_brent),
                frequency = frequency(obs_brent))

plot(obs_brent, type = "l", col = "black",
     main = "Observed vs Fitted: Brent",
     ylab = "Differenced Log Brent", xlab = "Time")
lines(fit_brent, col = "red")
legend("topright",
       legend = c("Observed", "Fitted"),
       col = c("black", "red"), lwd = 1)
```

```{r}
plot(obs_brent, fit_brent,
     pch = 16,
     xlab = "Observed",
     ylab = "Fitted",
     main = "Observed vs Fitted: Brent")
legend("topleft", 
       legend = paste("Correlation:", 
                      round(cor(obs_brent, fit_brent), 3)), 
       bty = "n")
abline(0, 1, col = "red", lwd = 2)
```

```{r}
# residual series
res_gas <- ts(res[, "Gasoline"],
              start = time(x_oil)[3],
              frequency = frequency(x_oil))

res_brent <- ts(res[, "Brent"],
                start = time(x_oil)[3],
                frequency = frequency(x_oil))

par(mfrow = c(2,2))

plot(res_gas, type = "l",
     main = "Residuals: Gasoline",
     ylab = "", xlab = "Time")
abline(h = 0, lty = 2, col = "gray")

acf(res[, "Gasoline"],
    main = "ACF: Gasoline Residuals")

plot(res_brent, type = "l",
     main = "Residuals: Brent",
     ylab = "", xlab = "Time")
abline(h = 0, lty = 2, col = "gray")

acf(res[, "Brent"],
    main = "ACF: Brent Residuals")
```

```{r}
# residual cross-correlation
par(mfrow = c(1,2))

ccf(res[, "Gasoline"], res[, "Brent"],
    main = "CCF: Gasoline and Brent Residuals")

ccf(res[, "Brent"], res[, "Gasoline"],
    main = "CCF: Brent and Gasoline Residuals")
```

```{r}
# normal Q-Q plots
par(mfrow = c(1,2))

qqnorm(res[, "Gasoline"],
       main = "Normal Q-Q Plot: Gasoline Residuals")
qqline(res[, "Gasoline"])

qqnorm(res[, "Brent"],
       main = "Normal Q-Q Plot: Brent Residuals")
qqline(res[, "Brent"])

par(mfrow = c(1,1))
```

Overall, the diagnostic plots suggest that the VAR(2) model provides an adequate fit to the differenced gasoline and crude oil series. The fitted values track the observed series reasonably well, particularly for gasoline, and the residuals fluctuate around zero without obvious trends or structure. The residual ACFs and CCFs indicate little remaining serial or cross-correlation beyond lag 0, suggesting that the model captures the main lead–lag relationships. Although the Q-Q plots show some deviations in the tails, especially for Brent, the residuals are otherwise reasonably well behaved, so there is no major evidence of model misspecification.

##### Questions: This doesn't seem like the best model for brent with gas as covariate, but is that relevant to us? since we care more about brent predicting gas.

### (v) Granger Causality to Further Investigate Direction

```{r}
causality(fit_var_oil, cause = "Brent")
causality(fit_var_oil, cause = "Gasoline")
```

The Granger causality tests indicate bidirectional predictive relationships between crude oil and gasoline prices. However, the predictive effect from oil to gasoline is substantially stronger than the reverse. This suggests that crude oil prices are the primary driver of gasoline price changes, while the predictive power of gasoline for oil likely reflects common shocks and market comovement rather than a strong causal effect.

### (vi) Forecast?
Don't know how to get 95% CI for fanchart

Plotted forecast using regular plot, fanchart, and ggplot; didn't know which style we want

```{r}
# pred_var_oil = predict(fit_var_oil, n.ahead = 24, ci = 0.95)
# plot(pred_var_oil)
# forecast(fit_var_oil, h = 24) |>
#   autoplot() + xlim(c(2020, 2027))
# fanchart(pred_var_oil, xlim = c(1500,1800))
```

```{r}
# 1 week
pred_var_oil_1 = predict(fit_var_oil, n.ahead = 1, ci = 0.95)
fc_oil_1 <- forecast(fit_var_oil, h = 1, level = 95)

# GAS
plot(pred_var_oil_1, names = "Gasoline",
     main = "Forecast for Gasoline",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_1, names = "Gasoline", 
#          main = "Forecast for Gasoline",
#          xlim = c(1500,1800))
# autoplot(fc_oil_1$forecast$Gasoline) +
#   ggtitle("Forecast for Gasoline") +
#   xlim(c(2020, 2027)) +
#   theme_minimal()

# BRENT
plot(pred_var_oil_1, names = "Brent",
     main = "Forecast for Brent",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_1, names = "Brent", 
#          main = "Forecast for Brent",
#          xlim = c(1500,1800))
# autoplot(fc_oil_1$forecast$Brent) +
#   ggtitle("Forecast for Brent") +
#   xlim(c(2020, 2027)) +
#   theme_minimal()
```

```{r}
# 4 weeks
forecasted <- predict(fit_var_oil, n.ahead = 4, ci = 0.95)

series_names <- c("Gasoline", "Brent")
titles <- c("Forecast for Gasoline (4 weeks)", "Forecast for Brent (4 weeks)")

for (i in 1:2) {
  name <- series_names[i]
  
  y_hist_full <- x_oil[, name]
  y_hist <- window(y_hist_full, start = 2025)
  t_hist_num <- as.numeric(time(y_hist))
  
  df <- as.data.frame(forecasted$fcst[[name]])
  t_fc <- tail(as.numeric(time(y_hist_full)), 1) + (1:4) / frequency(y_hist_full)
  
  # zoom in so the forecast is easier to see
  plot_start <- tail(t_hist_num, 26)[1]   # last ~26 weeks
  
  plot(y = y_hist,
       x = t_hist_num,
       type = "l",
       xlim = c(plot_start, max(t_fc)),
       ylim = range(c(y_hist, df$fcst, df$lower, df$upper), na.rm = TRUE),
       xlab = "Time",
       ylab = "",
       main = titles[i],
       xaxt = "n")
  
  tick_pos <- pretty(c(plot_start, max(t_fc)), n = 6)
  axis(1,
       at = tick_pos,
       labels = format(
         as.Date((tick_pos - 1970) * 365.25, origin = "1970-01-01"),
         "%m-%Y"
       ))
  
  abline(v = tail(as.numeric(time(y_hist_full)), 1), lty = 2, col = "gray")
  
  # connect forecast lines to the last observed point
  last_t <- tail(t_hist_num, 1)
  last_y <- tail(y_hist, 1)
  
  lines(x = c(last_t, t_fc),
        y = c(last_y, df$fcst),
        col = "blue", lty = 2)
  
  lines(x = c(last_t, t_fc),
        y = c(last_y, df$lower),
        col = "red", lty = 3)
  
  lines(x = c(last_t, t_fc),
        y = c(last_y, df$upper),
        col = "red", lty = 3)
}
```


```{r}
# 12 weeks
pred_var_oil_12 = predict(fit_var_oil, n.ahead = 12, ci = 0.95)
fc_oil_12 <- forecast(fit_var_oil, h = 12, level = 95)

# GAS
plot(pred_var_oil_12, names = "Gasoline",
     main = "Forecast for Gasoline",
     xlim = c(1500,1800))
fanchart(pred_var_oil_12, names = "Gasoline", 
         main = "Forecast for Gasoline",
         xlim = c(1500,1800))
autoplot(fc_oil_12$forecast$Gasoline) +
  ggtitle("Forecast for Gasoline") +
  xlim(c(2020, 2027)) +
  theme_minimal()

# BRENT
plot(pred_var_oil_12, names = "Brent",
     main = "Forecast for Brent",
     xlim = c(1500,1800))
fanchart(pred_var_oil_12, names = "Brent", 
         main = "Forecast for Brent",
         xlim = c(1500,1800))
autoplot(fc_oil_12$forecast$Brent) +
  ggtitle("Forecast for Brent") +
  xlim(c(2020, 2027)) +
  theme_minimal()
```

```{r}
# 13 weeks (Quarter of a year)
pred_var_oil_13 = predict(fit_var_oil, n.ahead = 13, ci = 0.95)
fc_oil_13 <- forecast(fit_var_oil, h = 13, level = 95)

# GAS
plot(pred_var_oil_13, names = "Gasoline",
     main = "Forecast for Gasoline",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_13, names = "Gasoline", 
#          main = "Forecast for Gasoline",
#          xlim = c(1500,1800))
# autoplot(fc_oil_13$forecast$Gasoline) +
#   ggtitle("Forecast for Gasoline") +
#   xlim(c(2020, 2027)) +
#   theme_minimal()

# BRENT
plot(pred_var_oil_13, names = "Brent",
     main = "Forecast for Brent",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_13, names = "Brent", 
#          main = "Forecast for Brent",
#          xlim = c(1500,1800))
# autoplot(fc_oil_13$forecast$Brent) +
#   ggtitle("Forecast for Brent") +
#   xlim(c(2020, 2027)) +
#   theme_minimal()
```

```{r}
# 24 weeks
pred_var_oil_24 = predict(fit_var_oil, n.ahead = 24, ci = 0.95)
fc_oil_24 <- forecast(fit_var_oil, h = 24, level = 95)

# GAS
plot(pred_var_oil_24, names = "Gasoline",
     main = "Forecast for Gasoline",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_24, names = "Gasoline", 
#          main = "Forecast for Gasoline",
#          xlim = c(1500,1800))
# autoplot(fc_oil_24$forecast$Gasoline) +
#   ggtitle("Forecast for Gasoline") +
#   xlim(c(2020, 2027)) +
#   theme_minimal()

# BRENT
plot(pred_var_oil_24, names = "Brent",
     main = "Forecast for Brent",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_24, names = "Brent", 
#          main = "Forecast for Brent",
#          xlim = c(1500,1800))
# autoplot(fc_oil_24$forecast$Brent) +
#   ggtitle("Forecast for Brent") +
#   xlim(c(2020, 2027)) +
#   theme_minimal()
```

```{r}
# 26 weeks (Half a year)
pred_var_oil_26 = predict(fit_var_oil, n.ahead = 26, ci = 0.95)
fc_oil_26 <- forecast(fit_var_oil, h = 26, level = 95)

# GAS
plot(pred_var_oil_26, names = "Gasoline",
     main = "Forecast for Gasoline (6 months)",
     xlim = c(1700,1750))
# fanchart(pred_var_oil_26, names = "Gasoline", 
#          main = "Forecast for Gasoline",
#          xlim = c(1500,1800))
# autoplot(fc_oil_26$forecast$Gasoline) +
#   ggtitle("Forecast for Gasoline") +
#   xlim(c(2020, 2027)) +
#   theme_minimal()

# BRENT
plot(pred_var_oil_26, names = "Brent",
     main = "Forecast for Brent (6 months)",
     xlim = c(1700,1750))
# fanchart(pred_var_oil_26, names = "Brent", 
#          main = "Forecast for Brent",
#          xlim = c(1500,1800))
# autoplot(fc_oil_26$forecast$Brent) +
#   ggtitle("Forecast for Brent") +
#   xlim(c(2020, 2027)) +
#   theme_minimal()
```

```{r}
# 36 weeks
pred_var_oil_36 = predict(fit_var_oil, n.ahead = 36, ci = 0.95)
fc_oil_36 <- forecast(fit_var_oil, h = 36, level = 95)

# GAS
plot(pred_var_oil_36, names = "Gasoline",
     main = "Forecast for Gasoline",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_36, names = "Gasoline", 
#          main = "Forecast for Gasoline",
#          xlim = c(1500,1800))
# autoplot(fc_oil_36$forecast$Gasoline) +
#   ggtitle("Forecast for Gasoline") +
#   xlim(c(2020, 2028)) +
#   theme_minimal()

# BRENT
plot(pred_var_oil_36, names = "Brent",
     main = "Forecast for Brent",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_36, names = "Brent", 
#          main = "Forecast for Brent",
#          xlim = c(1500,1800))
# autoplot(fc_oil_36$forecast$Brent) +
#   ggtitle("Forecast for Brent") +
#   xlim(c(2020, 2028)) +
#   theme_minimal()
```

```{r}
# 39 weeks (3/4 of a year)
pred_var_oil_39 = predict(fit_var_oil, n.ahead = 39, ci = 0.95)
fc_oil_39 <- forecast(fit_var_oil, h = 39, level = 95)

# GAS
plot(pred_var_oil_39, names = "Gasoline",
     main = "Forecast for Gasoline",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_39, names = "Gasoline", 
#          main = "Forecast for Gasoline",
#          xlim = c(1500,1800))
# autoplot(fc_oil_39$forecast$Gasoline) +
#   ggtitle("Forecast for Gasoline") +
#   xlim(c(2020, 2028)) +
#   theme_minimal()

# BRENT
plot(pred_var_oil_39, names = "Brent",
     main = "Forecast for Brent",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_39, names = "Brent", 
#          main = "Forecast for Brent",
#          xlim = c(1500,1800))
# autoplot(fc_oil_39$forecast$Brent) +
#   ggtitle("Forecast for Brent") +
#   xlim(c(2020, 2028)) +
#   theme_minimal()
```

```{r}
# 48 weeks
pred_var_oil_48 = predict(fit_var_oil, n.ahead = 48, ci = 0.95)
fc_oil_48 <- forecast(fit_var_oil, h = 48, level = 95)

# GAS
plot(pred_var_oil_48, names = "Gasoline",
     main = "Forecast for Gasoline",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_48, names = "Gasoline", 
#          main = "Forecast for Gasoline",
#          xlim = c(1500,1800))
# autoplot(fc_oil_48$forecast$Gasoline) +
#   ggtitle("Forecast for Gasoline") +
#   xlim(c(2020, 2028)) +
#   theme_minimal()

# BRENT
plot(pred_var_oil_48, names = "Brent",
     main = "Forecast for Brent",
     xlim = c(1500,1800))
# fanchart(pred_var_oil_48, names = "Brent", 
#          main = "Forecast for Brent",
#          xlim = c(1500,1800))
# autoplot(fc_oil_48$forecast$Brent) +
#   ggtitle("Forecast for Brent") +
#   xlim(c(2020, 2028)) +
#   theme_minimal()
```

```{r}
# 52 weeks (A year)
pred_var_oil_52 = predict(fit_var_oil, n.ahead = 52, ci = 0.95)
fc_oil_52 <- forecast(fit_var_oil, h = 52, level = 95)

# GAS
plot(pred_var_oil_52, names = "Gasoline",
     main = "Forecast for Gasoline (1 year)",
     xlim = c(1600,1800))
# fanchart(pred_var_oil_52, names = "Gasoline", 
#          main = "Forecast for Gasoline",
#          xlim = c(1500,1800))
# autoplot(fc_oil_52$forecast$Gasoline) +
#   ggtitle("Forecast for Gasoline") +
#   xlim(c(2020, 2028)) +
#   theme_minimal()

# BRENT
plot(pred_var_oil_52, names = "Brent",
     main = "Forecast for Brent (1 year)",
     xlim = c(1600,1800))
# fanchart(pred_var_oil_52, names = "Brent", 
#          main = "Forecast for Brent",
#          xlim = c(1500,1800))
# autoplot(fc_oil_52$forecast$Brent) +
#   ggtitle("Forecast for Brent") +
#   xlim(c(2020, 2028)) +
#   theme_minimal()
```


##### Question: how long do we want to forecast?

## 4. US Gasoline Prices Regional Comparison

### (i) Converting to Stationary Series

```{r}
# First-difference the log prices for all 5 regions
dRegions <- diff(log(regions_ts))
colnames(dRegions) <- colnames(regions_ts)

for (nm in colnames(dRegions)) {
  plot(dRegions[, nm],
       main = paste("Differenced Log", gsub("_", " ", nm)),
       ylab = "", xlab = "Time")
  
  acf(dRegions[, nm],
      lag.max = 20,
      main = paste("ACF:", gsub("_", " ", nm)))
  
  pacf(dRegions[, nm],
       lag.max = 20,
       main = paste("PACF:", gsub("_", " ", nm)))
}

par(mfrow = c(1, 1))

```

The differenced log models all seem stationary. From the plots of the differenced log models, they seem to be mean zero with no heteroskedasticity and no linear trend. The ACFs tend to decay quickly to zero, and the PACFs only have a spike close lag 0. There is no strong evidence of nonstationarity.

### (ii) CCF to understand lag dependence

```{r}
# Pairwise CCFs for all 10 regional pairs
region_pairs <- combn(colnames(dRegions), 2, simplify = FALSE)

# Put pairs involving Midwest first
region_pairs <- c(
  Filter(function(x) "Midwest" %in% x, region_pairs),
  Filter(function(x) !("Midwest" %in% x), region_pairs)
)

# Reorder within each pair so Midwest is first
region_pairs <- lapply(region_pairs, function(pair) {
  if ("Midwest" %in% pair) {
    c("Midwest", setdiff(pair, "Midwest"))
  } else {
    pair
  }
})

for (pair in region_pairs) {
  ccf(dRegions[, pair[1]], dRegions[, pair[2]],
      lag.max = 12,
      main = paste(gsub("_", " ", pair[1]), "vs", gsub("_", " ", pair[2])))
}

par(mfrow = c(1, 1))

```

Most of the regions are highly correlated with shared shocks (i.e. largest spike at zero). It's notable that the Midwest does tend to slightly lead the other regions because for the Midwest vs. \[region\] plots, while the largest spike is at zero, there is a stronger skew to the negative.

### (iii) VAR(1) Model

```{r}

x_regions <- dRegions

# Choose lag length
VARselect(x_regions, lag.max = 8, type = "const")

```

Tried 1, 3 and 8 but 1 gave best diagnostics.

```{r}
fit_var_regions <- VAR(x_regions, p = 1, type = "const")
summary(fit_var_regions)

```

Since all roots are less than 1, the model is stable.

**East Coast Model:** The East Coast gasoline prices have significant coefficients for one lagged difference of Midwest, Gulf Coast and West Coast. The Gulf Coast and Midwest coefficients have the relatively largest coefficients. It's notable that this model does not have a significant coefficient with its own lag.

**Midwest Model:** The Midwest gasoline prices have significant coefficients for one lagged difference with itself, East Coast, Gulf Coast and Rocky Mountain. Midwest and East Coast have comparable and the largest coefficients in terms of magnitude, but East Coast has a negative relationship. In addition, it's notable that this model has a significant relation with all regions except the West Coast.

**Gulf Coast:** The Gulf Coast gasoline prices has significant coefficients for one lagged difference of the Midwest, Rocky Mountain (though barely significant and small magnitude) and West Coast. It notably doesn't have a relationship with its own lag, but has the largest coefficient with the Midwest.

**Rocky Mountains:** All coefficients of the Rocky Mountain gasoline prices are significant, with the largest coefficient from the West Coast coefficient.

**West Coast:** All coefficients of the West Coast gasoline prices are significant, with the largest coefficient from its own lag.

For the correlation pattern of residuals, higher correlation means they are affected by the same shocks, whereas lower correlation means they are affected by more local dynamics. The East Coast, Midwest and Gulf Coast residuals are all strongly correlated (correlation higher than 0.7), which makes sense since they are adjacent regions and are affected by the same shocks. While the Rocky Mountain and West Coast are also correlated with the other regions, their correlations for residuals are never higher than 0.6, implying that they are likely affected by more local dynamics.

### (iv) Diagnostics

```{r}
# Fitted values and residuals
fitted_regions <- fitted(fit_var_regions)
res_regions <- resid(fit_var_regions)
```

```{r}
# Observed vs fitted plots for each region

p_used <- fit_var_regions$p

for (nm in colnames(x_regions)) {
  obs <- window(x_regions[, nm], start = time(x_regions)[p_used + 1])
  fit <- ts(fitted_regions[, nm],
            start = start(obs),
            frequency = frequency(obs))
  
  plot(obs, type = "l", col = "black",
       main = paste("Observed vs Fitted:", gsub("_", " ", nm)),
       ylab = "", xlab = "Time")
  lines(fit, col = "red")
  
  plot(obs, fit,
       pch = 16,
       xlab = "Observed",
       ylab = "Fitted",
       main = gsub("_", " ", nm))
  abline(0, 1, col = "red", lwd = 2)
  legend("topleft", 
       legend = paste("Correlation:", 
                      round(cor(obs, fit), 3)),
       bty = "n")

}

par(mfrow = c(1, 1))

```

Does good job everywhere except pre 2000s for rocky mountain. Midwest captures mean structure but no shocks.

```{r}
# Residual series and residual ACFs

for (nm in colnames(res_regions)) {
  plot(ts(res_regions[, nm],
          start = time(x_regions)[p_used + 1],
          frequency = frequency(x_regions)),
       type = "l",
       main = paste("Residuals:", gsub("_", " ", nm)),
       ylab = "", xlab = "Time", ylim = c(-0.2,0.2))
  abline(h = 0, lty = 2, col = "gray")
  
  acf(res_regions[, nm],
      main = paste("ACF Residuals:", gsub("_", " ", nm)))
}

par(mfrow = c(1, 1))

```

Residuals are largely white noise except Rocky Mountain has a slight decay in ACF.
again rocky mountain weird

```{r, fig.height = 6}
ts.plot(ts(res_regions[,1], 
           start = time(x_regions)[p_used + 1],
           frequency = frequency(x_regions)), 
        ts(res_regions[,2], 
           start = time(x_regions)[p_used + 1],
           frequency = frequency(x_regions)), 
        ts(res_regions[,3], 
           start = time(x_regions)[p_used + 1],
           frequency = frequency(x_regions)), 
        ts(res_regions[,4], 
           start = time(x_regions)[p_used + 1],
           frequency = frequency(x_regions)), 
        ts(res_regions[,5], 
           start = time(x_regions)[p_used + 1],
           frequency = frequency(x_regions)),
        col = 1:5, lty = 1,
        main = "Overlayed Residuals",
        xlab = "Year", ylab = "Dollars per Gallon")
legend("topleft",
       legend = c("East Coast", "Midwest", "Gulf Coast", "Rocky Mountain", "West Coast"),
       col = 1:5, lty = 1, bty = "n")

```


```{r}
resids_and_time <- data.frame(
  time = time(x_regions)[-1],
  avg_resid = rowMeans(res_regions),
  res_regions)
# highest average residual
resids_and_time |>
  dplyr::select(time, avg_resid) |>
  arrange(desc(avg_resid)) |> 
  head(5) |>
  arrange(time)
# highest East Coast residual
resids_and_time |>
  dplyr::select(time, East_Coast) |>
  arrange(desc(East_Coast)) |> 
  head(5) |>
  arrange(time)

# highest Midwest residual
resids_and_time |>
  dplyr::select(time, Midwest) |>
  arrange(desc(Midwest)) |> 
  head(5) |>
  arrange(time)

# highest Gulf Coast residual
resids_and_time |>
  dplyr::select(time, Gulf_Coast) |>
  arrange(desc(Gulf_Coast)) |> 
  head(5) |>
  arrange(time)

# highest Rocky Mountain residual
resids_and_time |>
  dplyr::select(time, Rocky_Mountain) |>
  arrange(desc(Rocky_Mountain)) |> 
  head(5) |>
  arrange(time)

# highest West Coast residual
resids_and_time |>
  dplyr::select(time, West_Coast) |>
  arrange(desc(West_Coast)) |> 
  head(5) |>
  arrange(time)
```

There is some overlap in the highest residuals for individual regions and overall average residuals.

```{r}
# Residual CCFs for all regional pairs
region_pairs <- combn(colnames(res_regions), 2, simplify = FALSE)

# Put pairs involving Rocky Mountain first
region_pairs <- c(
  Filter(function(x) "Rocky_Mountain" %in% x, region_pairs),
  Filter(function(x) !("Rocky_Mountain" %in% x), region_pairs)
)

# Reorder within each pair so Rocky Mountain is first
region_pairs <- lapply(region_pairs, function(pair) {
  if ("Rocky_Mountain" %in% pair) {
    c("Rocky_Mountain", setdiff(pair, "Rocky_Mountain"))
  } else {
    pair
  }
})

for (pair in region_pairs) {
  ccf(res_regions[, pair[1]], res_regions[, pair[2]],
      lag.max = 12,
      main = paste("Residual CCF:",
                   gsub("_", " ", pair[1]), "vs", gsub("_", " ", pair[2])))
}

par(mfrow = c(1, 1))

```

CCF is strange for Rocky Mountain comparison, but only has spikes at lag 0 for all other CCF comparisons.

**FINISH: why????**

```{r}
# Normal Q-Q plots

for (nm in colnames(res_regions)) {
  qqnorm(res_regions[, nm],
         main = paste("Q-Q:", gsub("_", " ", nm)))
  qqline(res_regions[, nm])
}

par(mfrow = c(1, 1))
```

Heavy tailed expected - maybe look into outliers? deviation from qq-line is likely to due larger shocks.

### (v) Granger Causality (pairwise)

Granger causality does not make sense for regional pairs, and any lags between regions can be determined by CCF and VAR coefficients.

### (vi) Forecast
Don't know how to get 95% CI for fanchart

Plotted forecast using regular plot, fanchart, and ggplot; didn't know which style we want

```{r}
# 1 week
pred_regions_1 <- predict(fit_var_regions, n.ahead = 1, ci = 0.95)

for (nm in colnames(x_regions)) {
  plot(pred_regions_1, 
       names = nm, 
       main = paste("Forecast for", gsub("_", " ", nm)),
       xlim = c(1500,1800))
}

# for (nm in colnames(x_regions)) {
#   fanchart(pred_regions_1, 
#            names = nm, 
#            main = paste("Forecast for", gsub("_", " ", nm)),
#            xlim = c(1500,1800))
# }
# 
# fc_regions_1 <- forecast(fit_var_regions, h = 1, level = 95)
# for (nm in colnames(x_regions)) {
#   p <- autoplot(fc_regions_1$forecast[[nm]]) +
#        ggtitle(paste("Forecast:", gsub("_", " ", nm))) +
#        xlim(c(2020, 2027)) +
#        theme_minimal()
#   print(p)
# }
```

```{r}
# 4 weeks (One month)
pred_regions_4 <- predict(fit_var_regions, n.ahead = 4, ci = 0.95)
for (nm in colnames(x_regions)) {
  plot(pred_regions_4, 
       names = nm, 
       main = paste("Forecast for", gsub("_", " ", nm), " (4 weeks)"),
       xlim = c(1700,1750))
}

# for (nm in colnames(x_regions)) {
#   fanchart(pred_regions_4, 
#            names = nm, 
#            main = paste("Forecast for", gsub("_", " ", nm)),
#            xlim = c(1500,1800))
# }
# 
fc_regions_4 <- forecast(fit_var_regions, h = 4, level = 95)
for (nm in colnames(x_regions)) {
  p <- autoplot(fc_regions_4$forecast[[nm]]) +
       ggtitle(paste("Forecast:", gsub("_", " ", nm))) +
       xlim(c(2020, 2027)) +
       theme_minimal()
  print(p)
}
```

```{r}
# 12 weeks
pred_regions_12 <- predict(fit_var_regions, n.ahead = 12, ci = 0.95)

for (nm in colnames(x_regions)) {
  plot(pred_regions_12, 
       names = nm, 
       main = paste("Forecast for", gsub("_", " ", nm)),
       xlim = c(1500,1800))
}

for (nm in colnames(x_regions)) {
  fanchart(pred_regions_12, 
           names = nm, 
           main = paste("Forecast for", gsub("_", " ", nm)),
           xlim = c(1500,1800))
}

fc_regions_12 <- forecast(fit_var_regions, h = 12, level = 95)
for (nm in colnames(x_regions)) {
  p <- autoplot(fc_regions_12$forecast[[nm]]) +
       ggtitle(paste("Forecast:", gsub("_", " ", nm))) +
       xlim(c(2020, 2027)) +
       theme_minimal()
  print(p)
}
```

```{r}
# 13 weeks (Quarter of a year)
pred_regions_13 <- predict(fit_var_regions, n.ahead = 13, ci = 0.95)

for (nm in colnames(x_regions)) {
  plot(pred_regions_13, 
       names = nm, 
       main = paste("Forecast for", gsub("_", " ", nm)),
       xlim = c(1500,1800))
}

# for (nm in colnames(x_regions)) {
#   fanchart(pred_regions_13, 
#            names = nm, 
#            main = paste("Forecast for", gsub("_", " ", nm)),
#            xlim = c(1500,1800))
# }
# 
# fc_regions_13 <- forecast(fit_var_regions, h = 13, level = 95)
# for (nm in colnames(x_regions)) {
#   p <- autoplot(fc_regions_13$forecast[[nm]]) +
#        ggtitle(paste("Forecast:", gsub("_", " ", nm))) +
#        xlim(c(2020, 2027)) +
#        theme_minimal()
#   print(p)
# }
```

```{r}
# 24 weeks
pred_regions_24 <- predict(fit_var_regions, n.ahead = 24, ci = 0.95)

for (nm in colnames(x_regions)) {
  plot(pred_regions_24, 
       names = nm, 
       main = paste("Forecast for", gsub("_", " ", nm)),
       xlim = c(1500,1800))
}

# for (nm in colnames(x_regions)) {
#   fanchart(pred_regions_24, 
#            names = nm, 
#            main = paste("Forecast for", gsub("_", " ", nm)),
#            xlim = c(1500,1800))
# }
# 
# fc_regions_24 <- forecast(fit_var_regions, h = 24, level = 95)
# for (nm in colnames(x_regions)) {
#   p <- autoplot(fc_regions_24$forecast[[nm]]) +
#        ggtitle(paste("Forecast:", gsub("_", " ", nm))) +
#        xlim(c(2020, 2027)) +
#        theme_minimal()
#   print(p)
# }
```

```{r}
# 24 weeks
pred_regions_24 <- predict(fit_var_regions, n.ahead = 24, ci = 0.95)

for (nm in colnames(x_regions)) {
  plot(pred_regions_24, 
       names = nm, 
       main = paste("Forecast for", gsub("_", " ", nm)),
       xlim = c(1500,1800))
}

# for (nm in colnames(x_regions)) {
#   fanchart(pred_regions_24, 
#            names = nm, 
#            main = paste("Forecast for", gsub("_", " ", nm)),
#            xlim = c(1500,1800))
# }
# 
# fc_regions_24 <- forecast(fit_var_regions, h = 24, level = 95)
# for (nm in colnames(x_regions)) {
#   p <- autoplot(fc_regions_24$forecast[[nm]]) +
#        ggtitle(paste("Forecast:", gsub("_", " ", nm))) +
#        xlim(c(2020, 2027)) +
#        theme_minimal()
#   print(p)
# }
```

```{r}
# 26 weeks (Half a year)
pred_regions_26 <- predict(fit_var_regions, n.ahead = 26, ci = 0.95)

for (nm in colnames(x_regions)) {
  plot(pred_regions_26, 
       names = nm, 
       main = paste("Forecast for", gsub("_", " ", nm), " (6 months)"),
       xlim = c(1700,1750))
}

# for (nm in colnames(x_regions)) {
#   fanchart(pred_regions_26, 
#            names = nm, 
#            main = paste("Forecast for", gsub("_", " ", nm)),
#            xlim = c(1500,1800))
# }
# 
# fc_regions_26 <- forecast(fit_var_regions, h = 26, level = 95)
# for (nm in colnames(x_regions)) {
#   p <- autoplot(fc_regions_26$forecast[[nm]]) +
#        ggtitle(paste("Forecast:", gsub("_", " ", nm))) +
#        xlim(c(2020, 2027)) +
#        theme_minimal()
#   print(p)
# }
```

```{r}
# 36 weeks
pred_regions_36 <- predict(fit_var_regions, n.ahead = 36, ci = 0.95)

for (nm in colnames(x_regions)) {
  plot(pred_regions_36, 
       names = nm, 
       main = paste("Forecast for", gsub("_", " ", nm)),
       xlim = c(1500,1800))
}

# for (nm in colnames(x_regions)) {
#   fanchart(pred_regions_36,
#            names = nm,
#            main = paste("Forecast for", gsub("_", " ", nm)),
#            xlim = c(1500,1800))
# }
# 
# fc_regions_36 <- forecast(fit_var_regions, h = 36, level = 95)
# for (nm in colnames(x_regions)) {
#   p <- autoplot(fc_regions_36$forecast[[nm]]) +
#        ggtitle(paste("Forecast:", gsub("_", " ", nm))) +
#        xlim(c(2020, 2028)) +
#        theme_minimal()
#   print(p)
# }
```

```{r}
# 39 weeks (3/4 of a year)
pred_regions_39 <- predict(fit_var_regions, n.ahead = 39, ci = 0.95)

for (nm in colnames(x_regions)) {
  plot(pred_regions_39, 
       names = nm, 
       main = paste("Forecast for", gsub("_", " ", nm)),
       xlim = c(1500,1800))
}

# for (nm in colnames(x_regions)) {
#   fanchart(pred_regions_39,
#            names = nm,
#            main = paste("Forecast for", gsub("_", " ", nm)),
#            xlim = c(1500,1800))
# }
# 
# fc_regions_39 <- forecast(fit_var_regions, h = 39, level = 95)
# for (nm in colnames(x_regions)) {
#   p <- autoplot(fc_regions_39$forecast[[nm]]) +
#        ggtitle(paste("Forecast:", gsub("_", " ", nm))) +
#        xlim(c(2020, 2028)) +
#        theme_minimal()
#   print(p)
# }
```

```{r}
# 48 weeks
pred_regions_48 <- predict(fit_var_regions, n.ahead = 48, ci = 0.95)

for (nm in colnames(x_regions)) {
  plot(pred_regions_48, 
       names = nm, 
       main = paste("Forecast for", gsub("_", " ", nm)),
       xlim = c(1500,1800))
}

# for (nm in colnames(x_regions)) {
#   fanchart(pred_regions_48, 
#            names = nm, 
#            main = paste("Forecast for", gsub("_", " ", nm)),
#            xlim = c(1500,1800))
# }
# 
# fc_regions_48 <- forecast(fit_var_regions, h = 48, level = 95)
# for (nm in colnames(x_regions)) {
#   p <- autoplot(fc_regions_48$forecast[[nm]]) +
#        ggtitle(paste("Forecast:", gsub("_", " ", nm))) +
#        xlim(c(2020, 2028)) +
#        theme_minimal()
#   print(p)
# }
```

```{r}
# 52 weeks (A year)
pred_regions_52 <- predict(fit_var_regions, n.ahead = 52, ci = 0.95)

for (nm in colnames(x_regions)) {
  plot(pred_regions_52, 
       names = nm, 
       main = paste("Forecast for", gsub("_", " ", nm), " (1 year)"),
       xlim = c(1600,1800))
}

# for (nm in colnames(x_regions)) {
#   fanchart(pred_regions_52, 
#            names = nm, 
#            main = paste("Forecast for", gsub("_", " ", nm)),
#            xlim = c(1500,1800))
# }
# 
# fc_regions_52 <- forecast(fit_var_regions, h = 52, level = 95)
# for (nm in colnames(x_regions)) {
#   p <- autoplot(fc_regions_52$forecast[[nm]]) +
#        ggtitle(paste("Forecast:", gsub("_", " ", nm))) +
#        xlim(c(2020, 2028)) +
#        theme_minimal()
#   print(p)
# }
```


