---
title: "36-671 Final Project"
author: "Ellie Huang, Nikita Jayaprakash, Annie Wang, Aishani Pradhan, Steve Wong"
date: "2026-04-19"
format: pdf
---

## Research Question

How do regional gasoline prices co-move over time, and what lead-lag relationships appear across regions and between crude oil prices and retail gasoline prices?

## 1. Preparing Time Series

```{r setup, include = FALSE}
# Core packages
library(here)
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(GGally)
library(vars)
library(forecast)
library(lubridate)
library(tibble)
```

```{r}
# File paths
gas_file_path <- here("data", "PET_PRI_GND_A_EPM0_PTE_DPGAL_W.xls")
brent_file_path <- here("data", "brent_crude_weekly_1993_present.csv")
```

```{r}
# Load weekly U.S. gasoline prices
us_gas_df <- read_excel(gas_file_path, sheet = "Data 1", skip = 2) |>
  rename(
    Date = 1,
    US_All_Grades = 2
  ) |>
  mutate(
    Date = as.Date(Date),
    US_All_Grades = as.numeric(US_All_Grades)
  ) |>
  arrange(Date)

# Load weekly regional gasoline prices
region_gas_df <- read_excel(gas_file_path, sheet = "Data 2", skip = 2) |>
  rename(
    Date = 1,
    East_Coast = 2,
    New_England = 3,
    Central_Atlantic = 4,
    Lower_Atlantic = 5,
    Midwest = 6,
    Gulf_Coast = 7,
    Rocky_Mountain = 8,
    West_Coast = 9,
    West_Coast_ex_CA = 10
  ) |>
  mutate(
    Date = as.Date(Date),
    across(-Date, as.numeric)
  ) |>
  arrange(Date)

# Keep the five regions used in the regional VAR
gas_df <- us_gas_df |>
  left_join(
    region_gas_df |>
      dplyr::select(Date, East_Coast, Midwest, Gulf_Coast, Rocky_Mountain, West_Coast),
    by = "Date"
  ) |>
  drop_na()

cat("Weekly gasoline date range:", min(gas_df$Date), "to", max(gas_df$Date), "\n")
cat("Gasoline observations:", nrow(gas_df), "\n")

# Load weekly Brent crude oil prices
brent_df <- read_csv(brent_file_path, show_col_types = FALSE) |>
  rename(Brent = Brent_Price_USD) |>
  mutate(
    Date = as.Date(Date),
    Brent = as.numeric(Brent)
  ) |>
  arrange(Date)

cat("Brent date range:", min(brent_df$Date), "to", max(brent_df$Date), "\n")
```

```{r}
# Helper to create weekly ts objects
make_weekly_ts <- function(x, start_date) {
  ts(
    x,
    start = c(
      as.numeric(format(start_date, "%Y")),
      as.numeric(format(start_date, "%U")) + 1
    ),
    frequency = 52
  )
}

gas_start_date <- min(gas_df$Date)

# Univariate gasoline series
us_gas_ts <- make_weekly_ts(gas_df$US_All_Grades, gas_start_date)
east_ts <- make_weekly_ts(gas_df$East_Coast, gas_start_date)
midwest_ts <- make_weekly_ts(gas_df$Midwest, gas_start_date)
gulf_ts <- make_weekly_ts(gas_df$Gulf_Coast, gas_start_date)
rocky_ts <- make_weekly_ts(gas_df$Rocky_Mountain, gas_start_date)
west_ts <- make_weekly_ts(gas_df$West_Coast, gas_start_date)

# Multivariate gasoline series for regional VAR analysis
region_var_ts <- gas_df |>
  dplyr::select(East_Coast, Midwest, Gulf_Coast, Rocky_Mountain, West_Coast) |>
  as.matrix() |>
  ts(
    start = c(
      as.numeric(format(gas_start_date, "%Y")),
      as.numeric(format(gas_start_date, "%U")) + 1
    ),
    frequency = 52
  )

colnames(region_var_ts) <- c(
  "East_Coast", "Midwest", "Gulf_Coast", "Rocky_Mountain", "West_Coast"
)
```

```{r}
# Align gasoline and Brent crude on common weekly dates
gas_brent_df <- gas_df |>
  inner_join(brent_df, by = "Date") |>
  drop_na()

cat(
  "Gasoline-Brent overlap:",
  format(min(gas_brent_df$Date), "%Y-%m-%d"), "to",
  format(max(gas_brent_df$Date), "%Y-%m-%d"), "\n"
)
cat("Overlap observations:", nrow(gas_brent_df), "\n")

gas_brent_start_date <- min(gas_brent_df$Date)

gas_ts_brent_overlap <- make_weekly_ts(gas_brent_df$US_All_Grades, gas_brent_start_date)
brent_ts <- make_weekly_ts(gas_brent_df$Brent, gas_brent_start_date)
```

## 2. Exploratory Data Analysis

```{r}
# National gasoline prices
plot(
  us_gas_ts,
  main = "U.S. All Grades Gasoline Prices (Weekly)",
  xlab = "Year",
  ylab = "Dollars per Gallon"
)

# Regional gasoline prices
ts.plot(
  east_ts, midwest_ts, gulf_ts, rocky_ts, west_ts,
  col = 1:5, lty = 1,
  main = "Regional Weekly Gasoline Prices",
  xlab = "Year",
  ylab = "Dollars per Gallon"
)
legend(
  "topleft",
  legend = c("East Coast", "Midwest", "Gulf Coast", "Rocky Mountain", "West Coast"),
  col = 1:5, lty = 1, bty = "n"
)

# Standardize gasoline and Brent for visual comparison
gas_scaled_ts <- ts(
  as.numeric(scale(as.numeric(gas_ts_brent_overlap))),
  start = start(gas_ts_brent_overlap),
  frequency = frequency(gas_ts_brent_overlap)
)

brent_scaled_ts <- ts(
  as.numeric(scale(as.numeric(brent_ts))),
  start = start(brent_ts),
  frequency = frequency(brent_ts)
)

ts.plot(
  brent_scaled_ts, gas_scaled_ts,
  col = c("steelblue", "firebrick"), lty = 1,
  main = "Standardized Brent Crude Oil and U.S. Gasoline Prices",
  xlab = "Year",
  ylab = "Standardized Value"
)
legend(
  "topleft",
  legend = c("Brent crude oil", "U.S. gasoline"),
  col = c("steelblue", "firebrick"),
  lty = 1,
  bty = "n"
)
```

The national gasoline series shows long swings and persistent level changes rather than fluctuations around a fixed mean. That pattern suggests the level series is nonstationary and should be transformed before lag-based modeling.

The five regional series move together closely during major price episodes, but they are not identical. The West Coast is usually the highest-priced region and appears more volatile, while the Gulf Coast and Midwest stay closer to the national pattern and tend to lie at lower price levels.

The standardized gasoline and Brent series also move closely together over time. This supports the idea that crude oil is an important input into retail gasoline pricing, while still leaving room for pass-through delays and region-specific effects.

```{r}
# Pairwise contemporaneous relationships across regions
gas_df |>
  dplyr::select(East_Coast, Midwest, Gulf_Coast, Rocky_Mountain, West_Coast) |>
  ggpairs(
    title = "Pairwise Regional Gasoline Prices",
    lower = list(continuous = wrap("points", alpha = 0.2, size = 0.5)),
    upper = list(continuous = wrap("cor", size = 3))
  )
```

The pairwise scatterplots show strong positive same-week relationships across all five regions. The reported correlations are uniformly high, ranging from 0.945 to 0.996, so the regions clearly are not evolving independently.

Several pairs, especially East Coast and Midwest, fall very tightly along a line. Pairs involving the West Coast are slightly more dispersed, which is consistent with the greater regional variation seen in the time-series plot.

These plots only describe contemporaneous association. They motivate a multivariate time-series approach, but they do not by themselves tell us which region moves first.

## 2.5. AR Model for U.S. Gasoline

```{r}
# Difference log gasoline prices to work with an approximately stationary series
d_us_gas <- diff(log(us_gas_ts))

# Hold out the last 4 weeks for a simple forecast check
forecast_horizon <- 4
n_gas <- length(d_us_gas)

train_d_us_gas <- window(d_us_gas, end = time(d_us_gas)[n_gas - forecast_horizon])
test_d_us_gas <- window(d_us_gas, start = time(d_us_gas)[n_gas - forecast_horizon + 1])

# Fit AR(2) model
fit_ar2_gas <- arima(train_d_us_gas, order = c(2, 0, 0))
summary(fit_ar2_gas)

# Forecast the held-out period
fc_ar2_gas <- forecast(fit_ar2_gas, h = forecast_horizon, level = 95)

fc_mean_ts <- ts(
  as.numeric(fc_ar2_gas$mean),
  start = start(test_d_us_gas),
  frequency = frequency(test_d_us_gas)
)

fc_lower_ts <- ts(
  as.numeric(fc_ar2_gas$lower[, 1]),
  start = start(test_d_us_gas),
  frequency = frequency(test_d_us_gas)
)

fc_upper_ts <- ts(
  as.numeric(fc_ar2_gas$upper[, 1]),
  start = start(test_d_us_gas),
  frequency = frequency(test_d_us_gas)
)

# Show recent history leading into the forecast origin
weeks_of_history <- 20
tail_start_time <- time(d_us_gas)[n_gas - forecast_horizon - weeks_of_history + 1]
actual_tail_ts <- window(d_us_gas, start = tail_start_time)

plot(
  actual_tail_ts,
  type = "l",
  lwd = 2,
  main = "AR(2) Forecast vs. Actual Near the Series End",
  xlab = "Time",
  ylab = "Differenced Log Gasoline",
  ylim = range(c(actual_tail_ts, fc_mean_ts, fc_lower_ts, fc_upper_ts))
)
lines(fc_mean_ts, lwd = 2, lty = 2, col = "red")
lines(fc_lower_ts, lty = 3, col = "blue")
lines(fc_upper_ts, lty = 3, col = "blue")
abline(v = time(test_d_us_gas)[1], lty = 2)
legend(
  "topleft",
  legend = c("Actual", "Forecast", "95% interval", "Forecast origin"),
  lty = c(1, 2, 3, 2),
  lwd = c(2, 2, 1, 1),
  bty = "n"
)

# Plot held-out actual values against forecasts
plot(
  test_d_us_gas,
  type = "l",
  lwd = 2,
  main = "Actual vs. Forecast: AR(2), 4-Week Holdout",
  xlab = "Time",
  ylab = "Differenced Log Gasoline",
  ylim = range(c(test_d_us_gas, fc_mean_ts, fc_lower_ts, fc_upper_ts))
)
lines(fc_mean_ts, lwd = 2, lty = 2, col = "red")
lines(fc_lower_ts, lty = 3, col = "blue")
lines(fc_upper_ts, lty = 3, col = "blue")
legend(
  "topleft",
  legend = c("Actual", "Forecast", "95% interval"),
  lty = c(1, 2, 3),
  lwd = c(2, 2, 1),
  bty = "n"
)
```

The AR(2) fit suggests that gasoline price changes are driven mostly by very recent history. The lag 1 coefficient is 0.5395, while the lag 2 coefficient is only 0.0696, so most of the predictable signal comes from the previous week rather than two weeks back.

The fitted mean is 0.0006, which is close to zero after differencing and logging. The residual lag 1 autocorrelation is -0.0049, so there is little obvious linear dependence left in the residuals.

The forecast plots show the main limitation of this univariate model. The 4-week forecasts quickly revert toward zero and miss the sharp movement at the end of the sample. Even though the observed values remain inside the forecast interval, the point forecasts are too smooth to track the sudden change. This suggests that a model based only on past gasoline prices is not enough when market conditions shift quickly.

## 3. Brent Crude Oil vs. U.S. Gasoline

### (i) Convert to approximately stationary series

```{r}
# First differences of log prices
d_us_gas_overlap <- diff(log(gas_ts_brent_overlap))
d_brent <- diff(log(brent_ts))

# Calendar dates corresponding to the differenced series
gas_brent_diff_dates <- gas_brent_df$Date[-1]

# Differenced log U.S. gasoline
plot(
  d_us_gas_overlap,
  main = "Differenced Log U.S. Gasoline",
  ylab = "Log difference",
  xlab = "Time"
)
abline(h = 0, lty = 2, col = "gray50")

par(mfrow = c(1, 2))
acf(d_us_gas_overlap, lag.max = 20,
  main = "ACF: Differenced Log U.S. Gasoline")

pacf(d_us_gas_overlap, lag.max = 20, main = "PACF: Differenced Log U.S. Gasoline")
par(mfrow = c(1, 1))

# Differenced log Brent crude
plot(
  d_brent,
  main = "Differenced Log Brent Crude",
  ylab = "Log difference",
  xlab = "Time"
)
abline(h = 0, lty = 2, col = "gray50")

par(mfrow = c(1, 2))
acf(d_brent, lag.max = 20, main = "ACF: Differenced Log Brent Crude")
pacf(d_brent, lag.max = 20, main = "PACF: Differenced Log Brent Crude")
par(mfrow = c(1, 1))
```

After differencing the log prices, both series fluctuate around a more stable mean and the autocorrelations decay much more quickly than in levels. The PACFs are dominated by the first few lags, which is consistent with using low-order dynamic models for the transformed series.

### (ii) CCF and lagged regression

```{r}
# Negative lag means Brent leads gasoline
ccf(
  d_brent, d_us_gas_overlap,
  lag.max = 12,
  main = "CCF: Brent vs. U.S. Gasoline (Differenced Log Prices)"
)
```

The cross-correlation function peaks at lag 0, which means gasoline and Brent often move strongly within the same week. There are also positive correlations at negative lags, especially around -1 to -3, which is consistent with Brent changes showing up in gasoline prices over the following weeks.

```{r}
# Build a lagged regression data set
oil_gas_lag_df <- ts.intersect(
  gas = d_us_gas_overlap,
  brent_lag_1 = stats::lag(d_brent, -1),
  brent_lag_2 = stats::lag(d_brent, -2),
  brent_lag_3 = stats::lag(d_brent, -3),
  dframe = TRUE
)

par(mfrow = c(1, 3))

plot(
  oil_gas_lag_df$brent_lag_1, oil_gas_lag_df$gas,
  xlab = "Brent change 1 week earlier",
  ylab = "Current gasoline change",
  main = "Lag 1"
)
abline(lm(gas ~ brent_lag_1, data = oil_gas_lag_df), col = "red")

plot(
  oil_gas_lag_df$brent_lag_2, oil_gas_lag_df$gas,
  xlab = "Brent change 2 weeks earlier",
  ylab = "Current gasoline change",
  main = "Lag 2"
)
abline(lm(gas ~ brent_lag_2, data = oil_gas_lag_df), col = "red")

plot(
  oil_gas_lag_df$brent_lag_3, oil_gas_lag_df$gas,
  xlab = "Brent change 3 weeks earlier",
  ylab = "Current gasoline change",
  main = "Lag 3"
)
abline(lm(gas ~ brent_lag_3, data = oil_gas_lag_df), col = "red")

par(mfrow = c(1, 1))
```

```{r}
# Regression using Brent lags 1 to 3
fit_oil_lag_lm <- lm(
  gas ~ brent_lag_1 + brent_lag_2 + brent_lag_3,
  data = oil_gas_lag_df
)

summary(fit_oil_lag_lm)
```

The lagged scatterplots support the CCF results. The association is strongest at a one-week lag, then becomes weaker at two and three weeks. In practical terms, this suggests that gasoline prices do not adjust all at once; some of the crude oil effect appears to pass through with a short delay.

### (iii) VAR(2) model

```{r}
# Bivariate stationary series for VAR analysis
oil_var_ts <- cbind(
  Gasoline = d_us_gas_overlap,
  Brent = d_brent
)

# Candidate lag lengths
VARselect(oil_var_ts, lag.max = 8, type = "none")
```

```{r}
# Fit the working VAR(2) model
fit_var_oil <- VAR(oil_var_ts, p = 2, type = "none")
summary(fit_var_oil)
```

We use a VAR(2) as a simple low-order model for the transformed gasoline and Brent series. This specification is consistent with the short-memory pattern in the ACF, PACF, and CCF plots and gives a direct way to estimate whether each series helps predict the other.

### (iv) Diagnostics

```{r}
# Fitted values and residuals
fitted_oil <- fitted(fit_var_oil)
resid_oil  <- resid(fit_var_oil)
```

```{r}
# Observed and fitted gasoline changes
obs_gas_oil <- window(oil_var_ts[, "Gasoline"], start = time(oil_var_ts)[3])
fit_gas_oil <- ts(
  fitted_oil[, "Gasoline"],
  start = start(obs_gas_oil),
  frequency = frequency(obs_gas_oil)
)

plot(
  obs_gas_oil, type = "l", col = "black",
  main = "Observed vs. Fitted: Gasoline",
  ylab = "Differenced Log Gasoline",
  xlab = "Time"
)
lines(fit_gas_oil, col = "red")
legend(
  "topright",
  legend = c("Observed", "Fitted"),
  col = c("black", "red"),
  lwd = 1,
  bty = "n"
)
```

```{r}
plot(
  obs_gas_oil, fit_gas_oil,
  pch = 16,
  xlab = "Observed",
  ylab = "Fitted",
  main = "Observed vs. Fitted: Gasoline"
)
legend(
  "topleft",
  legend = paste("Correlation:", round(cor(obs_gas_oil, fit_gas_oil), 3)),
  bty = "n"
)
abline(0, 1, col = "red", lwd = 2)
```

```{r}
# Largest absolute residuals for gasoline
resid_gas_oil <- resid_oil[, "Gasoline"]
oil_resid_time <- time(oil_var_ts)[-c(1, 2)]

top_resid_gas_idx <- order(abs(resid_gas_oil), decreasing = TRUE)[1:10]
top_resid_gas_dates <- oil_resid_time[top_resid_gas_idx] |>
  date_decimal() |>
  as.Date()

tibble(
  Date = top_resid_gas_dates,
  Gasoline_Residual = resid_gas_oil[top_resid_gas_idx]
)
```

```{r}
# Largest absolute residuals for Brent
resid_brent_oil <- resid_oil[, "Brent"]

top_resid_brent_idx <- order(abs(resid_brent_oil), decreasing = TRUE)[1:10]
top_resid_brent_dates <- oil_resid_time[top_resid_brent_idx] |>
  date_decimal() |>
  as.Date()

tibble(
  Date = top_resid_brent_dates,
  Brent_Residual = resid_brent_oil[top_resid_brent_idx]
)
```

```{r}
# Observed and fitted Brent changes
obs_brent_oil <- window(oil_var_ts[, "Brent"], start = time(oil_var_ts)[3])
fit_brent_oil <- ts(
  fitted_oil[, "Brent"],
  start = start(obs_brent_oil),
  frequency = frequency(obs_brent_oil)
)

plot(
  obs_brent_oil, type = "l", col = "black",
  main = "Observed vs. Fitted: Brent",
  ylab = "Differenced Log Brent",
  xlab = "Time"
)
lines(fit_brent_oil, col = "red")
legend(
  "topright",
  legend = c("Observed", "Fitted"),
  col = c("black", "red"),
  lwd = 1,
  bty = "n"
)
```

```{r}
plot(
  obs_brent_oil, fit_brent_oil,
  pch = 16,
  xlab = "Observed",
  ylab = "Fitted",
  main = "Observed vs. Fitted: Brent"
)
legend(
  "topleft",
  legend = paste("Correlation:", round(cor(obs_brent_oil, fit_brent_oil), 3)),
  bty = "n"
)
abline(0, 1, col = "red", lwd = 2)
```

```{r, fig.width = 8, fig.height = 6}
# Residual series
resid_gas_ts <- ts(
  resid_oil[, "Gasoline"],
  start = time(oil_var_ts)[3],
  frequency = frequency(oil_var_ts)
)

resid_brent_ts <- ts(
  resid_oil[, "Brent"],
  start = time(oil_var_ts)[3],
  frequency = frequency(oil_var_ts)
)

par(mfrow = c(2, 2))

plot(resid_gas_ts, type = "l", main = "Residuals: Gasoline", ylab = "", xlab = "Time")
abline(h = 0, lty = 2, col = "gray")
acf(resid_oil[, "Gasoline"], main = "ACF: Gasoline Residuals")

plot(resid_brent_ts, type = "l", main = "Residuals: Brent", ylab = "", xlab = "Time")
abline(h = 0, lty = 2, col = "gray")
acf(resid_oil[, "Brent"], main = "ACF: Brent Residuals")

par(mfrow = c(1, 1))
```

```{r}
# Residual cross-correlation
par(mfrow = c(1, 2))

ccf(resid_oil[, "Gasoline"], resid_oil[, "Brent"],
    main = "CCF: Gasoline and Brent Residuals")

ccf(resid_oil[, "Brent"], resid_oil[, "Gasoline"],
    main = "CCF: Brent and Gasoline Residuals")

par(mfrow = c(1, 1))
```

```{r, fig.width = 5, fig.height = 3}
# Normal Q-Q plots
par(mfrow = c(1, 2))

qqnorm(resid_oil[, "Gasoline"], main = "Normal Q-Q Plot: Gasoline Residuals")
qqline(resid_oil[, "Gasoline"])

qqnorm(resid_oil[, "Brent"], main = "Normal Q-Q Plot: Brent Residuals")
qqline(resid_oil[, "Brent"])

par(mfrow = c(1, 1))
```

Overall, the VAR(2) captures the broad short-run dependence reasonably well, especially for gasoline. The fitted-versus-observed plots show that the model tracks the center of the series, but it still misses the largest shocks, which is what we would expect from a low-order linear model fit to differenced prices.

The residual series fluctuate around zero and the residual ACFs show little remaining serial structure. The main departures appear in the tails of the Q-Q plots, especially for Brent, which suggests that unusual market weeks are still harder to explain than ordinary weeks.

### (v) Granger causality

```{r}
causality(fit_var_oil, cause = "Brent")
causality(fit_var_oil, cause = "Gasoline")
```

The Granger tests are best interpreted as predictive tests, not as proof of structural causation. In this context, bidirectional significance would mean that each series contains useful lagged information about the other, while a stronger Brent-to-gasoline result would still be more consistent with the economic story that crude oil price movements pass through into retail gasoline prices.

### (vi) 4-week holdout forecast

```{r, fig.width = 5, fig.height = 3}
# Hold out the last 4 weeks
forecast_horizon <- 4

oil_train_ts <- oil_var_ts[1:(nrow(oil_var_ts) - forecast_horizon), ]
oil_test_ts  <- oil_var_ts[(nrow(oil_var_ts) - forecast_horizon + 1):nrow(oil_var_ts), ]

oil_train_dates <- gas_brent_diff_dates[1:(length(gas_brent_diff_dates) - forecast_horizon)]
oil_test_dates  <- gas_brent_diff_dates[(length(gas_brent_diff_dates) - forecast_horizon + 1):length(gas_brent_diff_dates)]

fit_var_oil_train <- VAR(oil_train_ts, p = 2, type = "none")
oil_forecast <- predict(fit_var_oil_train, n.ahead = forecast_horizon, ci = 0.95)

series_names <- c("Gasoline", "Brent")
series_titles <- c(
  "Gasoline: Actual vs. Forecast (4-Week Test Set)",
  "Brent: Actual vs. Forecast (4-Week Test Set)"
)

for (i in seq_along(series_names)) {
  series_name <- series_names[i]

  y_train <- oil_train_ts[, series_name]
  y_test  <- oil_test_ts[, series_name]
  fc_df   <- as.data.frame(oil_forecast$fcst[[series_name]])

  plot_start_idx <- max(1, length(oil_train_dates) - 26 + 1)
  recent_dates <- oil_train_dates[plot_start_idx:length(oil_train_dates)]
  recent_train_values <- y_train[plot_start_idx:length(y_train)]

  plot(
    x = recent_dates,
    y = recent_train_values,
    type = "l",
    xlim = c(min(recent_dates), max(oil_test_dates)),
    ylim = range(c(y_train, y_test, fc_df$fcst, fc_df$lower, fc_df$upper), na.rm = TRUE),
    xlab = "Time",
    ylab = "",
    main = series_titles[i],
    xaxt = "n"
  )

  tick_pos <- seq(from = min(recent_dates), to = max(oil_test_dates), by = "1 month")
  axis(1, at = tick_pos, labels = format(tick_pos, "%m-%Y"))

  abline(v = tail(oil_train_dates, 1), lty = 2, col = "gray")

  lines(
    x = c(tail(oil_train_dates, 1), oil_test_dates),
    y = c(tail(y_train, 1), y_test),
    col = "black",
    lwd = 2
  )

  lines(
    x = c(tail(oil_train_dates, 1), oil_test_dates),
    y = c(tail(y_train, 1), fc_df$fcst),
    col = "blue",
    lty = 2,
    lwd = 2
  )

  lines(
    x = c(tail(oil_train_dates, 1), oil_test_dates),
    y = c(tail(y_train, 1), fc_df$lower),
    col = "red",
    lty = 3
  )

  lines(
    x = c(tail(oil_train_dates, 1), oil_test_dates),
    y = c(tail(y_train, 1), fc_df$upper),
    col = "red",
    lty = 3
  )

  legend(
    "topleft",
    legend = c("Train", "Actual test", "Forecast", "95% interval"),
    col = c("black", "black", "blue", "red"),
    lty = c(1, 1, 2, 3),
    lwd = c(1, 2, 2, 1),
    bty = "n"
  )
}
```

The 4-week holdout forecast shows that the bivariate VAR is still too smooth to capture the sharp spike at the end of the sample. Relative to the univariate AR model, it uses more information, but it still behaves like a short-memory model for differenced prices and therefore pulls quickly back toward zero.

That result matters for interpretation. The oil-gasoline link is useful for understanding dependence and pass-through, but it does not automatically produce strong short-horizon forecasts during abrupt market shifts. Forecast accuracy would likely improve only with additional information on broader market shocks, refining conditions, or policy disruptions.

## 4. Regional U.S. Gasoline Comparison

### (i) Convert regional series to approximately stationary form

```{r, fig.width = 5, fig.height = 3}
# First differences of log prices for all five regions
d_regions <- diff(log(region_var_ts))
colnames(d_regions) <- colnames(region_var_ts)

region_diff_dates <- gas_df$Date[-1]

for (series_name in colnames(d_regions)) {
  plot(d_regions[, series_name],
       main = paste("Differenced Log", gsub("_", " ", series_name)),
       ylab = "Log difference",
       xlab = "Time")
  abline(h = 0, lty = 2, col = "gray50")

  par(mfrow = c(1, 2))
  acf(d_regions[, series_name],
      lag.max = 20,
      main = paste("ACF:", gsub("_", " ", series_name)))
  pacf(d_regions[, series_name],
       lag.max = 20,
       main = paste("PACF:", gsub("_", " ", series_name)))
  par(mfrow = c(1, 1))
}
```

The transformed regional series appear much closer to stationarity than the level series. After differencing and logging, each series fluctuates around zero, the ACFs decay quickly, and there is no strong visual evidence of a remaining trend.

### (ii) CCFs for regional lead-lag patterns

```{r, fig.height = 12, fig.width = 10}
# Pairwise CCFs for all 10 regional pairs
region_pairs <- combn(colnames(d_regions), 2, simplify = FALSE)

# Show Midwest pairs first because they are especially informative in the plots
region_pairs <- c(
  Filter(function(x) "Midwest" %in% x, region_pairs),
  Filter(function(x) !("Midwest" %in% x), region_pairs)
)

region_pairs <- lapply(region_pairs, function(pair) {
  if ("Midwest" %in% pair) {
    c("Midwest", setdiff(pair, "Midwest"))
  } else {
    pair
  }
})

par(mfrow = c(5, 2))

for (pair in region_pairs) {
  ccf(d_regions[, pair[1]],
      d_regions[, pair[2]],
      lag.max = 12,
      main = paste(gsub("_", " ", pair[1]), "vs.", gsub("_", " ", pair[2])))
}

par(mfrow = c(1, 1))
```

Most regional pairs have their largest spike at lag 0, which indicates strong shared same-week movement. The Midwest comparisons also show some asymmetry at nearby negative lags, which is consistent with the Midwest sometimes moving slightly earlier than other regions, although the dominant pattern is still contemporaneous co-movement.

### (iii) VAR(1) model

```{r}
region_model_ts <- d_regions

# Candidate lag lengths
VARselect(region_model_ts, lag.max = 8, type = "const")
```

```{r}
# Fit the working regional VAR(1)
fit_var_regions <- VAR(region_model_ts, p = 1, type = "const")
summary(fit_var_regions)
roots(fit_var_regions)
```

We use a VAR(1) with an intercept as a parsimonious model for the five transformed regional series. The model is stable if all inverse roots lie inside the unit circle, and this specification is consistent with the short-memory behavior seen in the univariate ACFs.

In the fitted equations, the main pattern is that regions do help predict one another, but the strength of the relationships differs across the system. The West Coast and Rocky Mountain series tend to show more distinct regional behavior, while the East Coast, Midwest, and Gulf Coast appear more tightly linked.

### (iv) Diagnostics

```{r}
# Fitted values and residuals
fitted_regions <- fitted(fit_var_regions)
resid_regions  <- resid(fit_var_regions)
p_used <- fit_var_regions$p
```

```{r, fig.width = 10, fig.height = 5}
# Observed vs. fitted plots for each region
for (series_name in colnames(region_model_ts)) {
  obs_ts <- window(region_model_ts[, series_name], start = time(region_model_ts)[p_used + 1])
  fit_ts <- ts(fitted_regions[, series_name],
               start = start(obs_ts),
               frequency = frequency(obs_ts))

  par(mfrow = c(1, 2))

  plot(obs_ts, type = "l", col = "black",
       main = paste("Observed vs. Fitted:", gsub("_", " ", series_name)),
       ylab = "",
       xlab = "Time")
  lines(fit_ts, col = "red")

  plot(obs_ts, fit_ts,
       pch = 16,
       xlab = "Observed",
       ylab = "Fitted",
       main = gsub("_", " ", series_name))
  abline(0, 1, col = "red", lwd = 2)
  legend("topleft",
         legend = paste("Correlation:", round(cor(obs_ts, fit_ts), 3)),
         bty = "n")

  par(mfrow = c(1, 1))
}
```

The fitted values capture the center of the regional series reasonably well, but they do not capture abrupt shocks. This is especially noticeable for Rocky Mountain, where the fit appears less stable in earlier periods than in the other regions.

```{r, fig.width = 8, fig.height = 4}
# Residual series and ACFs
for (series_name in colnames(resid_regions)) {
  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

  plot(ts(resid_regions[, series_name],
          start = time(region_model_ts)[p_used + 1],
          frequency = frequency(region_model_ts)),
       type = "l",
       main = paste("Residuals:", gsub("_", " ", series_name)),
       ylab = "",
       xlab = "Time",
       ylim = c(-0.2, 0.2))
  abline(h = 0, lty = 2, col = "gray")

  acf(resid_regions[, series_name],
      main = paste("ACF Residuals:", gsub("_", " ", series_name)))

  par(mfrow = c(1, 1))
}
```

For most regions, the residuals look close to white noise. Rocky Mountain again stands out somewhat, with mild remaining structure relative to the rest of the system.

```{r, fig.width = 8, fig.height = 6}
# Overlay residuals from all regional equations
ts.plot(
  ts(resid_regions[, 1], start = time(region_model_ts)[p_used + 1], frequency = frequency(region_model_ts)),
  ts(resid_regions[, 2], start = time(region_model_ts)[p_used + 1], frequency = frequency(region_model_ts)),
  ts(resid_regions[, 3], start = time(region_model_ts)[p_used + 1], frequency = frequency(region_model_ts)),
  ts(resid_regions[, 4], start = time(region_model_ts)[p_used + 1], frequency = frequency(region_model_ts)),
  ts(resid_regions[, 5], start = time(region_model_ts)[p_used + 1], frequency = frequency(region_model_ts)),
  col = 1:5, lty = 1,
  main = "Overlayed Regional Residuals",
  xlab = "Year",
  ylab = "Differenced Log Price"
)
legend(
  "topleft",
  legend = c("East Coast", "Midwest", "Gulf Coast", "Rocky Mountain", "West Coast"),
  col = 1:5, lty = 1, bty = "n"
)
```

This overlay shows that large residuals tend to cluster during the same broad periods, which is consistent with system-wide shocks affecting several regions at once.

```{r}
# Weeks with the largest average and region-specific residuals
region_resid_df <- data.frame(
  time = time(region_model_ts)[-1],
  avg_resid = rowMeans(resid_regions),
  resid_regions
)

region_resid_df |>
  dplyr::select(time, avg_resid) |>
  arrange(desc(avg_resid)) |>
  head(5) |>
  arrange(time)

region_resid_df |>
  dplyr::select(time, East_Coast) |>
  arrange(desc(East_Coast)) |>
  head(5) |>
  arrange(time)

region_resid_df |>
  dplyr::select(time, Midwest) |>
  arrange(desc(Midwest)) |>
  head(5) |>
  arrange(time)

region_resid_df |>
  dplyr::select(time, Gulf_Coast) |>
  arrange(desc(Gulf_Coast)) |>
  head(5) |>
  arrange(time)

region_resid_df |>
  dplyr::select(time, Rocky_Mountain) |>
  arrange(desc(Rocky_Mountain)) |>
  head(5) |>
  arrange(time)

region_resid_df |>
  dplyr::select(time, West_Coast) |>
  arrange(desc(West_Coast)) |>
  head(5) |>
  arrange(time)
```

The largest residual weeks overlap across several regions, which again suggests that the biggest misses are driven by common market shocks rather than by isolated local noise.

```{r, fig.height = 12, fig.width = 10}
# Residual CCFs for all regional pairs
region_resid_pairs <- combn(colnames(resid_regions), 2, simplify = FALSE)

region_resid_pairs <- c(
  Filter(function(x) "Rocky_Mountain" %in% x, region_resid_pairs),
  Filter(function(x) !("Rocky_Mountain" %in% x), region_resid_pairs)
)

region_resid_pairs <- lapply(region_resid_pairs, function(pair) {
  if ("Rocky_Mountain" %in% pair) {
    c("Rocky_Mountain", setdiff(pair, "Rocky_Mountain"))
  } else {
    pair
  }
})

par(mfrow = c(5, 2))

for (pair in region_resid_pairs) {
  ccf(resid_regions[, pair[1]],
      resid_regions[, pair[2]],
      lag.max = 12,
      main = paste("Residual CCF:",
                   gsub("_", " ", pair[1]), "vs.", gsub("_", " ", pair[2])))
}

par(mfrow = c(1, 1))
```

For most pairs, the residual cross-correlations are concentrated at lag 0, which suggests the VAR(1) has removed most of the lagged dependence. Rocky Mountain still looks a bit less regular than the other regions, which is consistent with the earlier diagnostic plots.

```{r, fig.width = 8, fig.height = 6}
# Normal Q-Q plots
par(mfrow = c(2, 3))

for (series_name in colnames(resid_regions)) {
  qqnorm(resid_regions[, series_name],
         main = paste("Q-Q:", gsub("_", " ", series_name)))
  qqline(resid_regions[, series_name])
}

par(mfrow = c(1, 1))
```

The Q-Q plots show heavier tails than a normal distribution, which is not surprising for weekly energy price changes. The main message is that the regional VAR captures ordinary week-to-week dynamics better than it captures large shock periods.

### (v) Forecast

```{r, fig.width = 5, fig.height = 3}
# Hold out the last 4 weeks
forecast_horizon <- 4

region_train_ts <- region_model_ts[1:(nrow(region_model_ts) - forecast_horizon), ]
region_test_ts  <- region_model_ts[(nrow(region_model_ts) - forecast_horizon + 1):nrow(region_model_ts), ]

region_train_dates <- region_diff_dates[1:(length(region_diff_dates) - forecast_horizon)]
region_test_dates  <- region_diff_dates[(length(region_diff_dates) - forecast_horizon + 1):length(region_diff_dates)]

fit_var_regions_train <- VAR(region_train_ts, p = 1, type = "const")
region_forecast <- predict(fit_var_regions_train, n.ahead = forecast_horizon, ci = 0.95)

for (series_name in colnames(region_train_ts)) {
  y_train <- region_train_ts[, series_name]
  y_test  <- region_test_ts[, series_name]
  fc_df   <- as.data.frame(region_forecast$fcst[[series_name]])

  plot_start_idx <- max(1, length(region_train_dates) - 26 + 1)
  recent_dates <- region_train_dates[plot_start_idx:length(region_train_dates)]
  recent_train_values <- y_train[plot_start_idx:length(y_train)]

  plot(
    x = recent_dates,
    y = recent_train_values,
    type = "l",
    xlim = c(min(recent_dates), max(region_test_dates)),
    ylim = range(c(y_train, y_test, fc_df$fcst, fc_df$lower, fc_df$upper), na.rm = TRUE),
    xlab = "Time",
    ylab = "",
    main = paste("Actual vs. Forecast:", gsub("_", " ", series_name), "(4-Week Test Set)"),
    xaxt = "n"
  )

  tick_pos <- seq(from = min(recent_dates), to = max(region_test_dates), by = "1 month")
  axis(1, at = tick_pos, labels = format(tick_pos, "%m-%Y"))

  abline(v = tail(region_train_dates, 1), lty = 2, col = "gray")

  lines(
    x = c(tail(region_train_dates, 1), region_test_dates),
    y = c(tail(y_train, 1), y_test),
    col = "black",
    lwd = 2
  )

  lines(
    x = c(tail(region_train_dates, 1), region_test_dates),
    y = c(tail(y_train, 1), fc_df$fcst),
    col = "blue",
    lty = 2,
    lwd = 2
  )

  lines(
    x = c(tail(region_train_dates, 1), region_test_dates),
    y = c(tail(y_train, 1), fc_df$lower),
    col = "red",
    lty = 3
  )

  lines(
    x = c(tail(region_train_dates, 1), region_test_dates),
    y = c(tail(y_train, 1), fc_df$upper),
    col = "red",
    lty = 3
  )

  legend(
    "topleft",
    legend = c("Train", "Actual test", "Forecast", "95% interval"),
    col = c("black", "black", "blue", "red"),
    lty = c(1, 1, 2, 3),
    lwd = c(1, 2, 2, 1),
    bty = "n"
  )
}
```

Across regions, the 4-week forecasts again capture the average direction better than the exact week-to-week movements. Some regions, especially East Coast and West Coast, appear to show at least part of the short rise before dropping back, while others flatten out too quickly.

This reinforces the same point seen earlier: the regional VAR is useful for describing dependence across markets, but it is still limited as a short-horizon forecasting tool when the system is hit by sudden shocks.
