---
title: "36-671 Final Project"
author: "Ellie Huang, Nikita Jayaprakash, Annie Wang, Aishani Pradhan, Steve Wong"
date: "2026-04-19"
output: pdf_document
---

Research question: How do regional gasoline prices co-move over time, and what are the lead-lag relationships between regions and between crude oil prices and retail gasoline prices?

# Finding, downloading and cleaning up relevant time series data

```{r setup, include = FALSE}
library(here)
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(GGally)
library(vars)
library(forecast)

file_path <- here("data", "PET_PRI_GND_A_EPM0_PTE_DPGAL_W.xls")
brent_path <- here("data", "brent_crude_weekly_1993_present.csv")
```

```{r import-and-clean, message = FALSE}
# Load weekly gasoline data
# skip = 2 because row 1 = title, row 2 = source key, row 3 = actual variable names
us_data <- read_excel(file_path, sheet = "Data 1", skip = 2) |>
  rename(Date = 1, US_All_Grades = 2) |>
  mutate(Date = as.Date(Date), US_All_Grades = as.numeric(US_All_Grades)) |>
  arrange(Date)

region_data <- read_excel(file_path, sheet = "Data 2", skip = 2) |>
  rename(
    Date = 1,
    East_Coast = 2,
    New_England = 3,
    Central_Atlantic = 4,
    Lower_Atlantic = 5,
    Midwest = 6,
    Gulf_Coast = 7,
    Rocky_Mountain = 8,
    West_Coast = 9,
    West_Coast_ex_CA = 10
  ) |>
  mutate(Date = as.Date(Date)) |>
  mutate(across(-Date, as.numeric)) |>
  arrange(Date)

# Join national + five regional series, drop any incomplete weeks
gas_complete <- us_data |>
  left_join(
    region_data |> dplyr::select(Date, East_Coast, Midwest, Gulf_Coast, Rocky_Mountain, West_Coast),
    by = "Date"
  ) |>
  drop_na()

cat("Weekly gasoline date range:", min(gas_complete$Date), "to", max(gas_complete$Date), "\n")
cat("Observations:", nrow(gas_complete), "\n")

# Load weekly Brent crude oil data
crude_data <- read_csv(brent_path) |>
  rename(Brent = Brent_Price_USD) |>
  mutate(Date = as.Date(Date), Brent = as.numeric(Brent)) |>
  arrange(Date)

cat("Crude oil date range:", min(crude_data$Date), "to", max(crude_data$Date), "\n")
```

```{r weekly-ts}
# Convert all series to weekly ts objects
# Get week number via %U (Sunday start); +1 to make it start from 1
start_vec <- c(
  as.numeric(format(min(gas_complete$Date), "%Y")),
  as.numeric(format(min(gas_complete$Date), "%U")) + 1
)

us_ts <- ts(gas_complete$US_All_Grades, start = start_vec, frequency = 52)
east_ts <- ts(gas_complete$East_Coast, start = start_vec, frequency = 52)
midwest_ts <- ts(gas_complete$Midwest, start = start_vec, frequency = 52)
gulf_ts <- ts(gas_complete$Gulf_Coast, start = start_vec, frequency = 52)
rocky_ts <- ts(gas_complete$Rocky_Mountain, start = start_vec, frequency = 52)
west_ts <- ts(gas_complete$West_Coast, start = start_vec, frequency = 52)

# Build multivariate ts for VAR analysis
regions_ts <- gas_complete |>
  dplyr::select(East_Coast, Midwest, Gulf_Coast, Rocky_Mountain, West_Coast) |>
  as.matrix() |>
  ts(start = start_vec, frequency = 52)

colnames(regions_ts) <- c("East_Coast", "Midwest", "Gulf_Coast", "Rocky_Mountain", "West_Coast")
```

```{r gas-brent-join}
# Both series are weekly, join directly on date
gas_crude <- gas_complete |>
  inner_join(crude_data, by = "Date") |>
  drop_na()

cat("Gas price-crude oil overlapping date range:", min(gas_crude$Date), "to", max(gas_crude$Date), "\n")
cat("Observations:", nrow(gas_crude), "\n")

# Shared start for weekly ts
start_crude <- c(
  as.numeric(format(min(gas_crude$Date), "%Y")),
  as.numeric(format(min(gas_crude$Date), "%U")) + 1
)

gas_ts_crude <- ts(gas_crude$US_All_Grades, start = start_crude, frequency = 52)
crude_ts <- ts(gas_crude$Brent, start = start_crude, frequency = 52)
```

# Exploratory Data Analysis

```{r}
# Time series plots: national, regional, and crude oil
plot(us_ts,
     main = "U.S. All Grades Gasoline Prices (Weekly)",
     xlab = "Year", ylab = "Dollars per Gallon")

ts.plot(east_ts, midwest_ts, gulf_ts, rocky_ts, west_ts,
        col = 1:5, lty = 1,
        main = "Regional Weekly Gasoline Prices",
        xlab = "Year", ylab = "Dollars per Gallon")
legend("topleft",
       legend = c("East Coast", "Midwest", "Gulf Coast", "Rocky Mountain", "West Coast"),
       col = 1:5, lty = 1, bty = "n")

# Standardize for comparison since crude oil and gas prices are on diff scales/units
crude_scaled <- ts(
  as.numeric(scale(as.numeric(crude_ts))),
  start = start(crude_ts),
  frequency = frequency(crude_ts)
)

gas_scaled <- ts(
  as.numeric(scale(as.numeric(gas_ts_crude))),
  start = start(gas_ts_crude),
  frequency = frequency(gas_ts_crude)
)

ts.plot(crude_scaled, gas_scaled,
        col = c("steelblue", "firebrick"), lty = 1,
        main = "Standardized Brent Crude Oil vs. U.S. Gasoline Prices (Weekly)",
        xlab = "Year", ylab = "Standardized Value")
legend("topleft",
       legend = c("Brent Crude Oil", "U.S. Gasoline"),
       col = c("steelblue", "firebrick"), lty = 1, bty = "n")
```

The national gasoline series shows long swings over time rather than fluctuating around a constant level, suggesting strong persistence and possible nonstationarity in the level series. There seems to be a general upward movement over the sample. The plots also suggest changing variability over time, with larger swings during some periods, especially around major shocks, which may indicate heteroskedasticity.

The regional plot shows that all five regions follow the same broad pattern, rising and falling during the same major episodes.

At the same time, the regions are not exactly the same. The West Coast is generally higher than the other regions and appears more volatile, especially in later years. Rocky Mountain also shows somewhat greater variability at certain points, while the Gulf Coast and Midwest tend to track the national pattern more closely and at lower price levels. So the data show both strong shared movement and meaningful regional differences.

Comparing Brent crude to U.S. gasoline, the two series move very closely together over time. This suggests that crude oil is an important driver of gasoline prices, although there are other factors like refining, transportation, taxes, regional conditions, etc.

```{r}
# Pairwise scatterplots to check contemporaneous correlations before looking at lags
gas_complete |>
  dplyr::select(East_Coast, Midwest, Gulf_Coast, Rocky_Mountain, West_Coast) |>
  ggpairs(
    title = "Pairwise Regional Gasoline Prices",
    lower = list(continuous = wrap("points", alpha = 0.2, size = 0.5)),
    upper = list(continuous = wrap("cor", size = 3))
  )
```

The pairwise scatterplots indicate strong positive contemporaneous relationships among weekly gasoline prices across all regions. For every regional pair, the points form a clear upward-sloping pattern, suggesting that when prices are relatively high in one region, they also tend to be relatively high in the other during the same week. This is consistent with the very high pairwise correlations, which range from 0.945 to 0.996. Several pairs, such as East Coast and Midwest, fall especially tightly along a line, indicating very strong same-week linear association. Pairs involving the West Coast appear somewhat more dispersed by comparison, though the relationship remains strongly positive. However, because these plots compare observations from the same week only, they do not provide evidence about lead-lag structure or whether price changes in one region tend to precede changes in another.

This shows that the regions are clearly not independent; therefore it is reasonable to study them as a connected multivariate time series. Having established strong contemporaneous correlation, we can next investigate whether some regions tend to lead or lag others over time. This provides a natural motivation for using lag-based analysis and vector autoregressive models.

```{r, fig.width = 8, fig.height = 7}
# Cross-correlation between region pairs
# negative lag k means x leads y by k weeks
region_pairs <- list(
  c("East_Coast", "West_Coast"),
  c("East_Coast", "Gulf_Coast"),
  c("Midwest", "West_Coast"),
  c("Gulf_Coast", "Rocky_Mountain")
)

par(mfrow = c(2, 2))
for (pair in region_pairs) {
  ccf(gas_complete[[pair[1]]], gas_complete[[pair[2]]],
      main = paste("CCF:", pair[1], "vs", pair[2]),
      lag.max = 26,  # up to ~6 months of weekly lags
      ylab = "CCF")
}
par(mfrow = c(1, 1))
```

```{r}
# Cross-correlation: Brent vs. national gasoline
# negative lags = Brent leads gas, positive lags = gas leads Brent
ccf(gas_crude$Brent, gas_crude$US_All_Grades,
    main = "CCF: Brent Crude vs. U.S. Gasoline (Weekly)",
    lag.max = 26,
    ylab = "CCF")
```

The cross-correlation plots show that the regional gasoline price series are strongly related not just in the same week, but across many nearby lags as well. The highest correlations occur around lag 0, which points to strong contemporaneous movement (consistent with results from pairwise scatterplots above). At the same time, the correlations stay very high across the full lag window, so the raw series do not show a clear lead-lag pattern where one region consistently moves first and another follows after a specific number of weeks. The same pattern appears for Brent crude and U.S. gasoline prices, suggesting strong overall co-movement but not a clean lag interpretation in levels.

So we definitely need to check stationarity and do lead-lag interpretation on transformed stationary series rather than raw data.

```{r, fig.width = 8, fig.height = 7}
# ACF and PACF for national gasoline and Brent to characterize autocorrelation before fitting VAR
par(mfrow = c(2, 2))
acf(gas_complete$US_All_Grades, main = "ACF: U.S. Gasoline (Weekly)", lag.max = 52)
pacf(gas_complete$US_All_Grades, main = "PACF: U.S. Gasoline (Weekly)", lag.max = 52)
acf(gas_crude$Brent, main = "ACF: Brent Crude (Weekly)", lag.max = 52)
pacf(gas_crude$Brent, main = "PACF: Brent Crude (Weekly)", lag.max = 52)
par(mfrow = c(1, 1))
```

The ACF and PACF plots show that both weekly U.S. gasoline and Brent crude prices are highly persistent over time. The ACF stays large and declines slowly, which suggests that prices are strongly related to their past values even many weeks back. The PACF has 2 significant spikes at lags 1 and 2 and cuts off after, which suggests AR structure? Doesn't looks like there is clear seasonality over the first year of weekly lags.

# Model Selection - Regional VAR Model

```{r}
x = ts(dplyr::select(gas_complete, East_Coast:West_Coast), 
       # fourteenth week of 1993
       start = c(1993, 14), frequency = 52)
VARselect(x, lag.max = 14, type = 'both')
```
Using p = 2 due to EDA ACF/PACF graphs + SC (same as BIC).

```{r}
fit_regional = VAR(x, p = 2, type = 'both')
summary(fit_regional)
plot(ts(resid(fit_regional), start = c(1993, 14), frequency = 52))
```


```{r, fig.height = 9}
acf(resid(fit_regional), 52)
pred_regional = predict(fit_regional, n.ahead = 24, ci = 0.95)
plot(pred_regional)
forecast(fit_regional, h = 12) |>
  autoplot() + xlim(c(2020, 2027))
fanchart(pred_regional, xlim = c(1500,1800))
```

# Model Selection - Crude + US Retail VAR Model

```{r}
x = ts(dplyr::select(gas_crude, US_All_Grades, Brent), 
       # fourteenth week of 1993
       start = c(1993, 14), frequency = 52)
VARselect(x, lag.max = 14, type = 'both')
```
Using p = 3 due to SC (same as BIC).

```{r}
fit_crude = VAR(x, p = 3, type = 'both')
summary(fit_crude)
plot(ts(resid(fit_crude), start = c(1993, 14), frequency = 52))
```


```{r, fig.height = 6}
acf(resid(fit_crude), 52)
pred_crude = predict(fit_crude, n.ahead = 24, ci = 0.95)
plot(pred_crude)
forecast(fit_crude, h = 12) |>
  autoplot() + xlim(c(2020, 2027))
fanchart(pred_crude, xlim = c(1500,1800))
```